Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-08-30 Manuscript.qmd
Next draft version

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.
bibliography: references.bib
---

@@

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@mackenzie2005]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To ensure that relevant articles which did not directly cite @mackenzie2003 were captured, a second query searched articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

@@

A total of 60 articles were included for this review. Based on the acceptance rates within each strata, an estimated 516 of the 1300 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).

![Estimated coverage of the articles included in this review. A total of 60 articles from 2000-2023 were processed; bars indicating the estimated number of qualifying articles remaining are based on the rejection rate of each strata and the number of remaining unread articles in the pool. The proportion of articles included from each strata were as follows: 2004-2007, 12.5%; 2008-2011, 21.3%; 2012-2015, 44.4%; 2016-2019, 32.4%; 2020-2023, 57.1%.](Figures/CoveragePlot.jpeg){#fig-coverage}

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre [@basset2023], while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranging from under one month to forty years, with a median of 7 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

@@

While most articles in our sample fit a model to a single species, 26 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models used various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicitly interactive models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxa is likely to be reduced for practicality's sake.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 38 articles fit models to only a single taxa, 20 fit independent models to multiple taxa, and 6 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 10 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 3.25 years and covering 98 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

The presence/absence data required for DOMs has come from a breadth of different sources and detection methods. In our sample 76% of articles used human observational data, 19% used live-trapping methods, and 5% used detections from camera trapping or acoustic monitoring. Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger sign. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location. For example, in our only application of a DOM for a marine species @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' treating poachers as species which uses ranger reports to create detection histories.

Notably, many of these studies use data which was not originally collected in a robust design framework for occupancy modelling. In these articles, authors manipulated their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which far predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care in estimating detection probability, which is generally expected to vary based on observation method; however, this can be reasonably achieved by including survey method as a covariate on detection probability [@warrier2020; @pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. Drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019]. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data. For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days. While this provides interesting insights of occupancy at extremely fine temporal scale, further research may be needed on how to determine appropriate season and survey durations with respect to research questions and the closure assumption.

@@

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we have recorded all covariates considered for each model in our sample. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data is collected, and the ways in which those covariates are statistically represented in DOMs. We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environmental itself. Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of different covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry (31% of articles) and connectivity (26% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models. Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails. Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species. Examples include models with barred owl presence as a covariate on spotted owl occupancy [@mangan2019] and models with bullfrog covariates on native amphibian occupancy [@rowe2019], each of which effectively incorporate species interactions without fitting explicitly multi-species models.

@@

  filter(!Covariate %in% c("DIST", "PROT")) |>

@@

    "HABT" ~ html("<b>Habitat type (land)</b>\n<i>Aspects of terrestrial habitat</i>"),
    "HABA" ~ html("<b>Habitat type (aquatic)</b>\n<i>Aspects of aquatic habitats</i>"),

@@

    "GEOM" ~ html("<b>Geometry</b>\n<i>Site geometry, such as patch size or shape</i>"),
    "CONN" ~ html("<b>Connectivity</b>\n<i>Connectivity to other sites<i/>"),
    "LACO" ~ html("<b>Land cover</b>\n<i>Land cover classes (excluding agriculture)</i>"),
    "OTHER" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate and weather covariates</i>"),

@@

    "HYDR" ~ html("<b>Hydrology</b>\n<i>Hydrological features like rivers or lakes</i>"),
    "SPAT" ~ html("<b>Spatial</b>\n<i>Spatial variation, such as latitude</i>"),
    "AGRI" ~ html("<b>Agricultural</b>\n<i>Relating to agricultural features</i>"),
    "DISA" ~ html("<b>Disturbance</b>\n<i>Natural disasters or disturbances</i>"),

@@

    "OBSE" ~ html("<b>Observation</b>\n<i>Details of the survey</i>"),

@@

@@

Covariate data for studies in our sample was either collected directly by researchers (19% of environmental covariate data), or derived from pre-existing remotely sensed datasets (81%); this of course varies depending on the category of covariate. Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy. These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data is unavailable. For projects interested in making these projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002]. 29% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.

In the standard DOM covariates are incorporated for each parameter via a logit link, meaning that the statistical relationship between covariates and model parameters is represented as a linear term unless otherwise specified. Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model. @austin2007 discussed the importance of modelling ecologically-realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships. However, linear models are quite flexible, and non-linear responses can be easily accommodated using polynomial terms and interactions between covariates. Despite this, in our sample just 38% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships. Interactions between covariates were similarly rare, with only 28% of studies considering at least one relationship between terms. The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy. Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves where supported by the data [@merow2013; @elith2008]. This emphasis on more complex responses in SDMs may be due to their typically larger range-wide scale which encompasses the full species niche, where environmental relations may be expected to be non-linear. However, as previously indicated DOMs have been frequently implemented at similarly large spatial scales where the same assumptions should exist and similar responses might be expected. Similar concerns exist for the relatively low level of interactions between covariates: while they should not necessarily be used liberally, where these exist their exclusion can influence model performance [@guisan2006].

@@

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as 1 covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

@@

To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference. Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012]. Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by the formula $(2^n)^p$; where a simple single-parameter may have have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations. This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested. Models in our sample include both Frequentist (n = 47) and Bayesian (n = 13), which differ considerably in their manner of implementation. A summary of modelling practices in DOMs is given in @tbl-modelling below.

@@

The largest differences between frequentist and Bayesian models were in their approaches to model selection. Where 91% of frequentist models performed some manner of model selection to determine covariate inclusions for their models, only 31% of Bayesian models did so, with the majority instead fitting a single model defined *a priori*. Methods used for model selection varied considerably both within and between modelling frameworks. For frequentist models, the most popular and conventional approach to model selection as used in 49% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. Most frequentist studies which did not follow this protocol instead used procedural model selection methods (40% of articles), where the structure for each model parameter is fit one-by-one. For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on. A smaller portion of frequentist studies (4%) used more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM. Across all frequentist implementations, 70% of articles conducted multi-model inference and model-averaging using AIC weights [@burnham2004].

Those Bayesian models which *did* perform model selection took distinctly different approaches, with largely idiosyncratic methods across these studies. While direct comparison of model fit is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data. Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model. @ahumada2013 takes a particularly interesting hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.

Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is surely no one-size-fit all approach which will be appropriate for all possible use-cases. However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs @brodie2020. Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can make inaccurate estimates and influence the inference which we make from our models. @morin2020's work raises different concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models. The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.

>>>>

  A key aim of this review is to explore how users of DOMs have incorporated covariates into their models to describe variation in occupancy through space and time. Broadly speaking, model complexity in DOMs tends to be low, with relatively few covariates considered for inclusion and most relationships represented as simple linear terms. Considerable variation also exists in the procedures used for covariate selection, and limited research has been conducted on how these choices may influence model performance and inference. Furthermore, only a fraction of articles appear to evaluate models (either by goodness-of-fit testing or prediction validation). Likewise, little guidance exists on how to approach this task in DOMs. These uncertainties in the modelling process should be key priorities for future research on DOMs given their increasing use in applied ecological research.

bibliography: references.bib
csl: apa.csl 
embed-resources: true
---

@@

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To capture any additiona relevant articles which did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

@@

A total of 92 articles were included for this review. Based on the acceptance rates within each strata and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).

![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata. The proportion of articles included from each strata were as follows: 12.5% from 2005-2008; 24.4% from 2008-2011; 42.6% from 2012-2015; 35.1% from 2016-2019; and 57.1% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranging from under one month to forty years, with a median of 6 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

@@

While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.

![A) Study areas: Data for a majority of studies were collected from study locations in the United States. The size of study areas was log-normally distributed, with the median study area falling between 1000 and 10000 square kilometres. \n B) Study taxa: Most species modelled in our sample were terrestrial vertebrates including birds, mammals, or herptiles. 55 articles fit models to only a single taxa, 30 fit independent models to multiple taxa, and 11 fit multi-species models with explicit interactions between taxa. \n C) Survey methods: Approaches to data collected were varied and included conventional surveys like point counts and transects as well as more modern methods like camera traps and acoustic monitors. 13 articles incorporated data from citizen science projects. Project scale was also variable, with the median study running for 6 years and covering 101 sites.](Figures/ReviewBox.png){#fig-StudyDetails}

The presence/absence data required for DOMs has come from a breadth of different sources and detection methods. In our sample 79% of articles used human observational data, 9% used live-trapping methods, and 14% used detections from camera traps. Several articles also use citizen-science data to build detection histories, including @zuckerberg2011's model using bird-feeder data and @warrier2020's using citizen reports of tiger signs. Using citizen-science data may require additional consideration of assumptions (particularly false-positives), as discussed in greater detail by @cruickshank2019. Within these broad categories of detection method there is additional diversity, with several studies using unique survey methods determining the presence of a species at a location. For example, in the only application of a DOM for a marine species in our sample, @pendleton2022 uses aerial transects broken up into grid cells to observe whale occupancy; and @marescot2020's fits a unique 'multi-species' model treating poachers and using ranger reports to create detection histories.

Notably, many of these studies use data which were not originally collected in a robust design framework for occupancy modelling. In these articles, authors formatted their data into a hierarchical format post-hoc using a variety of methods. Some defined primary seasons as a discrete time interval, treating all surveys occurring within the season as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. These manipulations permit authors to use data which far predate the DOM, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

Additionally, not all articles rely on a single source of detection data - some integrate multiple sources of data to maximise sample size, combining data from camera traps, sign surveys, and citizen science reports. These integrated detection method models do require additional care and consideration; users must ensure that different detection methods represent comparable spatial and temporal scales, and that any variation in perceptibility between detection methods is accounted for [@warrier2020; @pitman2017]. A special case exists when different detection methods are used where one has the potential for false positive detections; e.g., where less-certain citizen science detections are combined with certain detections from field surveys. In this context, the certain detections are used to help account for false-positive detection probability, as in @miller2011's study integrating GPS collars and hunter reports to estimate wolf occupancy in Montana.

The flexibility in the data used for DOMs, including the model structure and the scale of observations, is not amenable to a one-size-fits-all definition of occupancy. Users of DOMs must carefully consider precisely what they are modelling and address questions on the scale represented by their model [@chave2013]. Intepretation and drivers of occupancy may differ depending on whether a site is represented by a single point on the landscape or as a grid cell, with the former likely to depend on more local, small scale factors rather than landscape-level trends [@stevens2019]. This is also true for the temporal scale of occupancy: whether a site is occupied within a week or within a year leads to vastly different conceptions of occupancy. This idea is particularly relevant in cases where the selection of season length *is* to some extent arbitrary, as with camera-trap or bioacoustic data where continuous recordings can be broken down into distinct 'seasons' of any length. DOMs are well suited to these data types [@balantic2019], and the proliferation of autonomous survey techniques provides novel opportunities for analysis that is simply not possible with human-collected data. For example, @kleiven2020 and @mölle2022 divide their camera trap records into seasons of just a few days. While this provides interesting insights of occupancy at extremely fine temporal scale, further research may be needed on how to determine appropriate season and survey durations with respect to research questions and the closure assumption.

@@

Building dynamic occupancy models can be a challenging process requiring careful consideration of which environmental factors to incorporate to adequately represent occupancy and detection in complex natural systems. In this review, we have recorded all covariates considered for each model in our sample. These are summarised in a taxonomy presented in @tbl-covariates, which states the proportion of articles considering each type of covariate in their models, the means by which that covariate data are collected, and how covariates responses are represented in DOMs. We further categorise these covariates into two groups: environmental covariates, which represent plausible ecological correlates of parameters; and structural covariates, representing aspects of model form and observation functionally distinct from the environment. Our findings indicate that users of DOMs have incorporated an exceptionally wide diversity of covariates in their models — while the most common varieties of covariates include aspects of habitat and land cover, a range of other unique factors were considered by authors in our sample. Many models incorporate covariates representing aspects of site geometry and connectivity (37% of articles), adding spatial components to these DOMs which bear similarities to related stochastic patch occupancy models. Notable examples of these spatially-explicit models include @mortelliti2007's model fit to rodent data in patchy landscapes and @risk2011's hybrid DOM/incidence function model used to explore meta-population dynamics of rails. Several studies also include biotic interactions with other species as covariates, particularly those which describe taxa threatened by invasive species. Examples include models with barred owl presence as a covariate on spotted owl occupancy and models with bullfrog covariates on native amphibian occupancy, each of which effectively incorporate species interactions without fitting explicitly multi-species models [@mangan2019; @rowe2019].

@@

@@

    "HABT" ~ html("<b>Habitat </b>\n<i>Aspects of habitat and land cover</i>"),

@@

    "SPAT" ~ html("<b>Spatial</b>\n<i>Site geometry, connectivity, or other spatial elements</i>"),
    "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography, including hydrology</i>"),
    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),

@@

@@

    "OBSE" ~ html("<b>Observation</b>\n<i>Details on the observation process</i>"),

@@

    "OTHER-S" ~ html("<b>Other structural</b>\n<i>Other structural covariate not otherwise listed</i>"),

@@

Covariate data for studies in our sample was either collected directly by researchers (17% of environmental covariate data), or derived from pre-existing remotely sensed datasets (83%); this of course varies depending on the category of covariate. Directly collected data can generally be expected to represent finer-scale factors like prey species occurrence or details of habitat structure, which can be difficult to measure remotely but can often be more proximal drivers of occupancy. These high quality covariates do come with trade-offs, as they can be expensive to collect and can often preclude projection to locations where data are unavailable. For projects interested in making these projections, remotely-sensed covariates are generally more feasible despite their generally more distal nature [@austin2002]. 42% of environmental factors and 87% of structural factors included in reviewed articles were dynamic covariates which varied through time — this again varied with the category of covariate in question, with terms relating to climate or weather most frequently dynamic and topographic covariates universally static.

In the standard DOM, covariates for each parameter are most commonly incorporated via a logistic regression (i.e., a linear regression through a logit link function). Statistical relationships between covariates and model parameters are represented as linear terms unless otherwise specified. Of course, many ecological relationships are non-linear and require more complex forms to be realistically represented in a model. @austin2007 discussed the importance of modelling ecologically-realistic responses to covariates, advocating for careful consideration of the most appropriate statistical form for hypothesised relationships. However, linear models are quite flexible, and non-linear responses can be easily accommodated using polynomial terms and interactions between covariates. Despite this, in our sample just 35% of articles considered using at least one non-linear responses to an environmental covariate, with the majority of studies representing all covariates as simple linear relationships. Interactions between covariates were similarly rare, with only 26% of studies considering at least one relationship between terms. The relatively low emphasis on more complex non-linear responses contrasts with other popular methods for modelling species occupancy. Many common approaches for SDMs, which include MAXENT and Boosted Regression Trees, permit considerable flexibility in the shape of their covariate response curves where supported by the data [@merow2013; @elith2008]. This emphasis on more complex responses in SDMs may be due to their typically larger range-wide scale which encompasses the full species niche, where environmental relations may be expected to be non-linear. However, as previously indicated DOMs have been frequently implemented at similarly large spatial scales where the same assumptions should exist and similar responses might be expected. Similar concerns exist for the relatively low level of interactions between covariates: while they should not necessarily be used liberally, where these exist their exclusion can influence model performance [@guisan2006].

@@

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents an alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as 1 covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

@@

To this point in our review, we have discussed only the covariates which were *considered* for inclusion in DOMs without regard for which terms were actually *included* in the final models used by authors to make their inference. Identifying the 'best' model from a range of possible candidates can be a challenging process, especially for hierarchical models like DOMs [@doherty2012]. Consider that the quantity of candidate models involving up to *n* covariates on *p* parameters can be given by $(2^n)^p$; where a simple single-parameter may have up to 64 possible combinations of 6 covariates, a 4-parameter dynamic occupancy model would have 16 million combinations. This explosion of candidate models generally precludes exhaustive comparisons of possible models and requires a reduction of the models tested. Models in our sample include both Frequentist (n = 70) and Bayesian (n = 22), which differ considerably in their manner of implementation. A summary of modelling practices in DOMs is given in @tbl-modelling below.

@@

The largest differences between the frequentist and Bayesian models in our sample lie in their approaches to model selection. Where 94% of frequentist models perform some manner of model selection to determine covariate inclusions for their models, only 36% of Bayesian models do so, with the majority instead fitting a single model defined *a priori*. Methods used for model selection vary considerably both within and between modelling frameworks. For frequentist models, the most popular and conventional approach to model selection as used in 43% of articles involves the creation of a fully pre-defined model set containing some number of hypothesis candidate models, where the best model is selected according to the lowest AIC score. Most frequentist studies which do not follow this protocol instead use procedural model selection methods (39% of articles), where the structure for each model parameter is fit one-by-one. For example, this protocol might first identify the best structure for detection probability before moving on to initial occupancy and so on. A smaller portion of frequentist studies (9%) use more unique approaches, fitting simpler models such as single season occupancy models for candidate covariates to identify the most informative terms to use in a DOM. Across all frequentist implementations, 71% of articles conduct multi-model inference and model-averaging using AIC weights [@burnham2004].

Those Bayesian models which *do* perform model selection tend to take distinctly different approaches, with largely idiosyncratic methods across these studies. While direct comparison of model fit is rare, it is feasible — @urban2023 identifies the best model from a Bayesian candidate set using the predictive performance on both in and out-of-sample validation data. Another approach used by @cook2022 fit a global model including all covariates before removing each covariate where the 95% credible interval of the posterior distribution overlapped zero and refitting the model. @ahumada2013 takes a hybrid approach, in which model selection is conducted by a procedural method in the frequentist framework before refitting the best structure in a Bayesian model.

Limited research has been conducted on advantages of different methods for covariate selection in DOMs, and there is surely no one-size-fit all approach which will be appropriate for all possible use-cases. However, it is important to consider the implications of the different model selection approaches in common usage, and research from the SDM literature highlights how covariate selection can influence our interpretation of model outputs [@brodie2020]. Within the occupancy modelling literature, @stewart2023's article on covariate selection in single-season occupancy model discusses important attributes of the information-criteria approaches widely used in frequentist models in our review, noting that these approaches can lead to inaccurate estimates and influence model inference. @morin2020's work raises different concerns on procedural model selection methods, demonstrating how the fine details of modelling protocols can determine whether the best-fitting possible model is identified and which covariates appear in final models. The literature on performance of Bayesian model selection methods is more sparse, although @hooten2015's guide to Bayesian model selection in ecology is a valuable resource for possible methods of fitting those models.

With few exceptions, most of the model selection methods identified in our review are designed only to identify the best model of those considered, and do not necessarily guarantee that selected models are well-fit to the data. Assessment of the performance of DOMs generally requires additional steps, and the best ways to achieve this are not necessarily clear. There is no broadly-accepted goodness-of-fit test for dynamic occupancy models, although @mackenzie2004 describes an approach for single-season occupancy models using a parametric bootstrap which can been extended to DOMs; this test is implemented in the *AICcModAvg* and *unmarked* R packages. @kéry2021 describe the test and present an alternative based on separately assessing fit to static and dynamic components of the model. In Bayesian implementations, posterior predictive checks offer means to assess model fit [@gelman2014]. @broms2016 discusses possibilities for model evaluation in the Bayesian context focusing on single-season multi-species occupancy models but extensions of their approach may also be applicable to the DOM. As in other hierarchical models, model evaluation for DOMs can in general can be difficult and somewhat uncertain compared to other model types, as the primary response variable of interest (species occupancy) is a latent variable where the true state is generally not known. Predictive performance evaluation is thus typically based on *observed* occupancy data, where a DOM is used to simulate detection histories to be compared with field survey results.
